---
title: "Client Report - Further Housing Data Analysis "
subtitle: "Housing and Student Living Office"
author: "Kayson Garn"
format:
  html:
    self-contained: true
    page-layout: full
    title-block-banner: true
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    html-math-method: katex
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
    code-copy: hover
    code-tools:
        source: false
        toggle: true
        caption: See code
execute: 
  warning: false
---

# Further Housing Analysis

```{r}
#| message: false
#| warning: false
#| include: false
#to load neccessary libraries
library(tidyverse)
library(mosaic)
library(pander)
library(dplyr)
library(stringr)
library(margins)
library(scales)
library(rio)
library(plotly)
```

```{r}
#to read in the data
HSL_data <- rio::import("HSL_data.csv")
```

```{r}
#to investigate what bins should be created...
HSL_data <- HSL_data |> 
  mutate(property_size  = case_when(
    Total_Beds < 20 ~ "House",
    Total_Beds >=20 & Total_Beds <80 ~ "Small",
    Total_Beds >=80 & Total_Beds <=180 ~ "Medium",
    Total_Beds >180 ~ "Large"
  ))


hist_bins <- ggplot(HSL_data, aes(x=as.factor(property_size))) + geom_bar() + theme_minimal()

hist_bins
```

So we define the following

**House:** **x** \< 20

**Small:** 20 \<= **x** \< 80

**Medium**: 80 \<= **x** \<= 180

**Large**: **x** \> 180

Lets Look at the differences in prices compared to the different sized apartment complexes.

### Original Price Comparison

```{r}
HSL_years_high_occ_grouped <- HSL_data |> 
  filter(!is.na(Cleanliness)) |> 
  group_by(Semester, Year, Full_Occ) |> 
  summarise(
    mean_price = mean(Mean_Contract_Price),
    median_occ = median(Occ_Rate) * 100,
    mean_beds = mean(Total_Beds),
    mean_management = mean(Management),
    mean_maintenance = mean(Maintenance_Response),
    mean_cleanliness = mean(Cleanliness),
    mean_activities = mean(Activities_Events),
    mean_culture = mean(Culture),
    mean_value = mean(Value),
    mean_overall = mean(Overall)
  ) |> 
  mutate(
    Semester = case_when(
      Semester == "SP" ~ "Spring",
      Semester == "WI" ~ "Winter"
    )
  )
```

```{r}
#| message: false
#| warning: false
price_chart <- ggplot(HSL_years_high_occ_grouped, 
                      aes(x = Year, y = mean_price, color = Full_Occ)) + 
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) + 
  facet_wrap(~Semester) +
  scale_color_manual(
    values = c(
      "TRUE" = "#006EB6",
      "FALSE" = "#949598" 
    )
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(size = 16, face = "bold"),
    panel.spacing = unit(1.5, "lines")
  ) + 
  labs(
    title = "Student Approved Housing 2022-2025",
    x = "Year",
    y = "Mean Price",
    color = "Full Occupancy"
  )

ggplotly(price_chart)
```

### Price Comparison by Property Size

```{r}
HSL_size_spring <- HSL_data |> 
  filter(!is.na(Cleanliness)) |> 
  group_by(Semester, Year, Full_Occ, property_size) |> 
  summarise(
    mean_price = mean(Mean_Contract_Price),
    median_occ = median(Occ_Rate) * 100,
    mean_beds = mean(Total_Beds),
    mean_management = mean(Management),
    mean_maintenance = mean(Maintenance_Response),
    mean_cleanliness = mean(Cleanliness),
    mean_activities = mean(Activities_Events),
    mean_culture = mean(Culture),
    mean_value = mean(Value),
    mean_overall = mean(Overall)
  ) |> 
  mutate(
    Semester = case_when(
      Semester == "SP" ~ "Spring",
      Semester == "WI" ~ "Winter"
    ) 
  ) |>
    filter(Semester == "Spring")
```

```{r, fig.height=10, fig.width=10}
price_chart_spring <- ggplot(HSL_size_spring, 
                      aes(x = Year, y = mean_price, color = Full_Occ)) + 
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) + 
  facet_wrap(~property_size) + 
  theme_minimal() + 
  labs(
    y= "Mean Price",
    x=" ",
    title="Spring Price Comparison"
  ) + 
  theme(
    legend.position = "bottom"
  )

price_chart_spring
```

```{r}
HSL_size_winter <- HSL_data |> 
  filter(!is.na(Cleanliness)) |> 
  group_by(Semester, Year, Full_Occ, property_size) |> 
  summarise(
    mean_price = mean(Mean_Contract_Price),
    median_occ = median(Occ_Rate) * 100,
    mean_beds = mean(Total_Beds),
    mean_management = mean(Management),
    mean_maintenance = mean(Maintenance_Response),
    mean_cleanliness = mean(Cleanliness),
    mean_activities = mean(Activities_Events),
    mean_culture = mean(Culture),
    mean_value = mean(Value),
    mean_overall = mean(Overall)
  ) |> 
  mutate(
    Semester = case_when(
      Semester == "SP" ~ "Spring",
      Semester == "WI" ~ "Winter"
    ) 
  ) |>
    filter(Semester == "Winter")
```

```{r, fig.height=10, fig.width=10}
price_chart_winter <- ggplot(HSL_size_winter, 
                      aes(x = Year, y = mean_price, color = Full_Occ)) + 
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) + 
  facet_wrap(~property_size) + 
  theme_minimal() + 
  labs(
    y= "Mean Price",
    x=" ",
    title="Spring Price Comparison"
  ) + 
  theme(
    legend.position = "bottom"
  )

price_chart_winter
```

### Original Occupancy Comparison

```{r}
#| message: false
#| warning: false
occ_chart <- ggplot(HSL_years_high_occ_grouped, aes(x=Year, y=median_occ, color= Full_Occ)) + 
  geom_line(size = 1.2) +
  geom_point(size = 3) + 
  facet_wrap(~Semester) +
  scale_color_manual(
    values = c(
      "TRUE" = "#006EB6",
      "FALSE" = "#949598" 
    )) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(size = 16, face = "bold"),
    panel.spacing = unit(1.5, "lines")
  ) + 
  labs(
    title = "Student Approved Housing 2022-2025",
    x = "Year",
    y = "Mean Occupancy",
    color = "Full Occupancy"
  )

ggplotly(occ_chart)
```

### Occupancy Comparison by Property Size

```{r, fig.height=10, fig.width=10}
occ_chart_spring <- ggplot(HSL_size_spring, 
                      aes(x = Year, y = median_occ, color = Full_Occ)) + 
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) + 
  facet_wrap(~property_size) + 
  theme_minimal() + 
  labs(
    y= "Mean Occupancy",
    x=" ",
    title="Spring Occupancy Comparison"
  ) + 
  theme(
    legend.position = "bottom"
  )

occ_chart_spring
```

```{r, fig.height=10, fig.width=10}
occ_chart_winter <- ggplot(HSL_size_winter, 
                      aes(x = Year, y = median_occ, color = Full_Occ)) + 
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) + 
  facet_wrap(~property_size) + 
  theme_minimal() + 
  labs(
    y= "Mean Occupancy",
    x=" ",
    title="Winter Occupancy Comparison"
  ) + 
  theme(
    legend.position = "bottom"
  )

occ_chart_winter
```

### Original Beds Comparison

```{r}
#| message: false
#| warning: false
bedcount_chart <- ggplot(HSL_years_high_occ_grouped, aes(x=Year, y=mean_beds, color= Full_Occ)) + 
  geom_line(size = 1.2) +
  geom_point(size = 3) + 
  facet_wrap(~Semester) +
  theme_minimal() +
  scale_color_manual(
    values = c(
      "TRUE" = "#006EB6",
      "FALSE" = "#949598" 
    )) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(size = 16, face = "bold"),
    panel.spacing = unit(1.5, "lines")
  ) + 
  labs(
    title = "Student Approved Housing 2022-2025",
    x = "Year",
    y = "Mean Total Beds",
    color = "Full Occupancy"
  )

ggplotly(bedcount_chart)
```

### Bed Count Comparison by Property Size

```{r, fig.height=10, fig.width=10}
bed_chart_spring <- ggplot(HSL_size_spring, 
                      aes(x = Year, y = mean_beds, color = Full_Occ)) + 
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) + 
  facet_wrap(~property_size) + 
  theme_minimal() + 
  labs(
    y= "Mean Bed Count",
    x=" ",
    title="Spring Bed Count Comparison"
  ) + 
  theme(
    legend.position = "bottom"
  )

bed_chart_spring
```

```{r, fig.height=10, fig.width=10}
bed_chart_winter <- ggplot(HSL_size_winter, 
                      aes(x = Year, y = mean_beds, color = Full_Occ)) + 
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) + 
  facet_wrap(~property_size) + 
  theme_minimal() + 
  labs(
    y= "Mean Bed Count",
    x=" ",
    title="Winter Bed Count Comparison"
  ) + 
  theme(
    legend.position = "bottom"
  )

bed_chart_winter
```

# Creating the New Variable

Here we want to determine the differences between properties that are full year-round, not-full year-round, full in winter but not spring, and full in spring but not winter.

We will start off by creating that variable titled "Year Occupancy Categories".

```{r}
#lets start off by pivoting the table wider so that each property has only one row per year:

HSL_wide <- HSL_data |> 
  select(Property, Year, Semester, boolocc) |> 
  pivot_wider(names_from = Semester, values_from = boolocc) |> 
  rename(occ_winter = WI, occ_spring = SP)
```

```{r}
#now we want to create the actual categorical variables: 

HSL_wide <- HSL_wide |> 
  mutate(Year_Occ_Category = case_when(
    occ_winter==1 & occ_spring == 1 ~"Full Year-Round",
    occ_winter==0 & occ_spring==0 ~"Not Full Year-Round",
    occ_winter==1 & occ_spring==0~ "Full Winter, Not Spring",
    occ_winter==0 & occ_spring==1 ~"Full Spring, Not Winter"
  ))
```

```{r}
#now I need to join this new variable onto the original dataset 
HSL_data <- HSL_data |> 
  left_join(HSL_wide |> select(Property, Year, Year_Occ_Category),
            by = c("Property","Year"))
```

```{r}
HSL_data |>
  group_by(Year_Occ_Category) |>
  summarize(
    avg_occ_rate = mean(Occ_Rate, na.rm = TRUE),
    avg_price    = mean(Mean_Contract_Price, na.rm = TRUE),
    avg_becount    = mean(Total_Beds, na.rm = TRUE),
    avg_maintenance  = mean(Maintenance_Response, na.rm = TRUE),
    avg_cleanliness  = mean(Cleanliness, na.rm = TRUE),
    avg_culture  = mean(Culture, na.rm = TRUE),
    n_properties = n_distinct(Property),
    avg_culture  = mean(Culture, na.rm = TRUE),
  )
```
